\documentclass{article}
\title{X16 Edit user manual}
\date{\today}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1.2ex}

\usepackage{hyperref}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{longtable}

\begin{document}

\begin{huge}
    X16 Edit user manual
\end{huge}

\vspace{1em}
\today

\vspace{4em}
\tableofcontents
\vspace{4em}

\section{Introduction}

Thank you for trying out X16 Edit!

X16 Edit is a simple text editor written in 65C02 assembly especially
for the Commander X16 retro computer. 

The look and feel of the program is inspired by GNU Nano. There are,
naturally, a lot of differences, but you will feel at home if you are
used to Nano.

The Commander X16 was devised by David Murray a.k.a. the 
8-Bit Guy. For more information on the platform, go to 
\href{http://www.commanderx16.com}{www.commanderx16.com}.

\section{Basic usage}
 
    \subsection{Getting started}
        As of Kernal version R45, the editor is stored in ROM and can be
        started with the BASIC command EDIT.

        There is also a RAM based version of the editor. It consists of just one file, 
        X16EDIT-x.x.x.PRG, where x.x.x is the version number. The RAM based editor 
        is loaded and run in the same way as a BASIC program.
        
        To run the RAM based editor on hardware, just store the program file
        on an SD card. Insert the SD card into your X16 computer. Type
        \texttt{LOAD"X16EDIT-x.x.x.PRG"} to load the program. And then \texttt{RUN} to start it.
 
        If you want to run the RAM based editor in the x16emu emulator, you first need to 
        store the X16 Edit executable in the host computer's file system.
        Then type the following command in the terminal:
 
\begin{verbatim}
    x16emu -prg X16EDIT-x.x.x.PRG -run
\end{verbatim}

        A valid path to the editor executable must be specified after
        the "-prg" option. Otherwise the emulator will not know where
        to find the program.
 
        The "-run" option automatically starts the editor after the emulator
        is booted. You may remove this option and start the editor manually
        by typing \texttt{RUN} in the emulator.
 
    \subsection{Entering text}
        X16 Edit is a modeless text editor. As soon as it is started, everything
        you type is stored in the text buffer.
 
        The cursor is moved with keys normally used for that purpose on a modern
        keyboard, \textit{i.e.} the arrow keys, Tab, Home, End, PgUp and PgDn.

        The go to line feature (Ctrl+L) lets you move the cursor to a specific line number.
 
        As you type and reach the right margin, the editor does not by default insert a line break. 
        Instead the current line is scrolled horizontally to
        make room for more characters. There is no limit to the length of a line
        other than the available memory.
 
    \subsection{Saving and loading text files}
        The current text buffer is saved to file on the SD card when you press Ctrl+O. Type in the 
        file name you like to use, and press the ENTER key. You will
        be prompted to confirm before overwriting a file that already exists.
         
        To load a file from the SD card, press Ctrl+R. 
        Just type in the name of the file you want to load, and then press the ENTER key.
        
    \subsection{Keyboard shortcuts}
        X16 Edit is controlled by keyboard shortcuts. You find a complete list
        of shortcuts in Appendix A.

        Shortcuts are primarily selected by pressing and holding down Ctrl at the
        same time as you press a shortcut key.

        You may also press and release the ESC key. A message is displayed
        in the status bar indicating that the program is ready to receive a command.
        Just press a shortcut key without holding down any modifier.

    \subsection{Mouse control}
        The editor is primarily made to be controlled by the keyboard. There are, however,
        some basic mouse controls.

        The cursor may be positioned in the file by clicking the left mouse button.

        You may select part of the text by pressing the left mouse button where
        you want the selection to start and release it where you want the selection to end. 
        Selections can be copied or cut. More on that later.

        The mouse is automatically enabled when you start the editor. The mouse pointer 
        is hidden while typing on the keyboard, and shown again when you move or click the mouse.
    
    \subsection{User interface}
        X16 Edit's user interface is similar to GNU Nano, and should be mostly self-explanatory.
        It consists of the following main parts:

        \begin{itemize}
            \item Title bar
            \item Editor area
            \item Status bar
            \item Shortcut list
        \end{itemize}

        The \textit{title bar} is at the top row of the screen. The name of the current file
        is displayed at the center. If the text buffer has not been saved to file, it displays
        "New text buffer". At the right edge the letters "MOD" are shown if the
        text buffer has been changed since last saved to file.

        The \textit{editor area} is right below the title bar. It takes up most of the screen, and
        it is here you do all text editing.

        The \textit{status bar} is at the third row from the bottom of the screen. All messages
        from the program are displayed in the status bar. If the program needs to prompt you
        for input, the input prompt is displayed here as well. 
        
        At the input prompt you may either press ENTER to confirm your input or press ESC 
        to abort the current operation.

        The active \textit{line break encoding} and the current \textit{cursor position} are displayed at the left and right edge of the status bar respectively, if
        there is no active message or prompt.

        The \textit{shortcut list} takes up the last two rows of the screen. Here you find
        the most common keyboard shortcuts.
        
        X16 Edit uses the screen mode (screen height and width) that is active when starting
        the program. Especially the shortcut list and the help screen change appearance to fit in 
        low screen resolutions.

    \subsection{Built-in help}
        A lists of all keyboard shortcuts and a short description 
        of each command is available in the built-in help screen (Ctrl+G). 
    
\section{More on text editing}

    \subsection{Supported character sets}
        The editor supports the following built-in character sets:

        \begin{itemize}
            \item PETSCII upper case/graphics (standard and thin). This is the default mode of both the Commander X16 and the C64.

            \item PETSCII upper/lower case (standard and thin). This is the same mode as is available on the C64.

            \item ISO character set (standard and thin). This mode is new for the X16, and there is no corresponding mode 
            supported by Commodore 8 bit computers. Text is encoded according to ISO-8859-15, making it
            easier to transfer files to and from modern computers.

            \item CP437.

            \item Cyrillic (standard and thin).

            \item Eastern latin (standard and thin).
            
        \end{itemize}

        On startup, the editor detects the current character set and continues using that.

        You may change the character set by pressing Ctrl+E. Press left or right cursor keys to
        browse the supported character sets, and press Enter to select the new character set you want to use.
    
    \subsection{Quote mode}
        The PETSCII character set contains control codes that are commonly used in BASIC programs. The
        control codes are, for instance, used to clear the screen and to move the cursor. Normally, it is not
        possible to type PETSCII control codes in the editor. Control codes may
        be typed by activating the quote mode (Ctrl+Q). You exit quote mode by pressing ESC.

    \subsection{Tab stop}
        The default tab stop width is four spaces. 
        
        You may change the width by Ctrl+1--9.
                
    \subsection{Auto-indent}
        The auto-indent feature is used to keep the level of indentation when line breaks are inserted manually or automatically by
        the word wrap feature.

        Auto-indent is turned off when the editor starts. To toggle the feature on or off, press Ctrl+A.

    \subsection{Word wrap}
        The editor does not automatically insert line breaks by default. Instead the current line is scrolled horizontally
        when you reach the right margin.

        Automatic word wrap may be toggled on or off by pressing Ctrl+Z. When turned on you are prompted for the column
        where to wrap.

        The word wrap feature works in a very simplified way. When you reach the right margin the editor breaks the
        line after the previous blank space. If there is no blank space on the line, the line break is inserted
        at the right margin. The line break position is not recalculated if you delete characters from the line
        or if you insert new characters at the beginning of the line.

    \subsection{Text justification}
        The justify command (Ctrl+J) divides the text buffer into paragraphs and recalculates all line breaks
        according to the line width specified in the word wrap feature (default 80). The word wrap function 
        need not be enabled to run the justify function.
          
        When auto-indent is turned off a new paragraph is considered to begin

        \begin{itemize}
            \item if two or more consecutive line breaks are found, or
            \item if a line starts with one or more blank spaces.
        \end{itemize}

        If auto-indent is turned on, a new paragraph is considered to begin

        \begin{itemize}
            \item if two or more consecutive line breaks are found, 
            \item if a line contains only blank space characters, or
            \item if the level of indentation is different from the previous line.
        \end{itemize}

    \subsection{Cut, copy and uncut}
        The editor supports cut, copy, and uncut (commonly called paste in other editors).

        Parts of the text can be selected for copying or cutting with the keyboard or mouse.
        A selection is made with the keyboard by holding down Shift while you move the cursor
        with the arrow keys or the Home or End keys. A selection can similarily be made by
        holding the left mouse button and move the pointer to the end of the selection before
        the button is released.

        If no text has been selected, the copy (Ctrl+C) and cut (Ctrl+K) commands
        copy a whole line into the clipboard.

        You may do multiple copies or cuts into the clipboard. The clipboard holds a maximum of 3 kB of data.

        Uncut (Ctrl+U) copies the content of the clipboard into the text buffer at the cursor position. It is possible
        to do repeated uncuts. The clipboard is cleared upon the first cut or copy after uncutting.

    \subsection{Search and replace}
        The search command (Ctrl+W) lets you search the text buffer for
        a particular string. The replace function (Ctrl+S)
        lets you replace it with another string.

        Both search and replace are case sensitive. Search starts from the character after the cursor
        and is forward looking.

        If the string you are searching for is found, the cursor is moved to that position.

        When replacing a string, you are given the option to only replace the
        next occurrence or all subsequent occurrences.

    \subsection{Line break encoding}
        The currently used line break encoding is displayed at the left edge of the status bar.

        The editor's default behavior is to use CR (ASCII 13) as line break when started in
        one of the PETSCII modes, and LF (ASCII 10) when started in ISO mode.

        On opening a file, the editor detects the use of any of the following line break encodings:

        \begin{itemize}
            \item CR -- Commodore or old Mac style
            \item LF -- Unix style
            \item CRLF -- Windows style
        \end{itemize}

        The line break encoding found in the file is thereafter used by the editor.

        You can manually change the active line break encoding by pressing Ctrl+Enter.

\section{More on file handling}
    \subsection{Commodore DOS file paths}
    The editor accepts a full Commodore DOS file path when prompting you for a file name.

    A Commodore DOS file path consists of a directory and a file name separated by a colon, for
    example:

    \begin{itemize}
        \item //MY-APP/:APPNAME.PRG
        \item /MY-APP/:APPNAME.PRG
    \end{itemize}
    
    The first example starting with double slashes is an absolute path from the root of the SD card. The second
    example starting with a single slash is a realtive path starting from the current directory.

    \subsection{The built-in file browser}
        At the prompt where a file name is entered you may alternatively press Ctrl+T to show the
        built-in file browser. The file browser is there to make it easier to move between directories 
        and select files.
        
        To select an item in the file browser, first highlight it with the up or down arrow keys, and then press Enter.

        If the selected item is a directory, it will be made the new current directory, and its content will
        be displayed in the file browser.

        If not all items fit on one page the listing is ended with "--- MORE ---". 
        In case the items are spread over several pages, you may go to 
        the next page with PgDn or Ctrl+V and back to a previous page with PgUp or Ctrl+Y.

        If there are no more items to show the listing is ended with "--- END ---".

    \subsection{Change disk drive device number}
        By default the file handling functions use device \#8. The device
        number may be changed by pressing Ctrl+D.

    \subsection{Disk drive commands}
        You may invoke disk drive commands by pressing Ctrl+I. The raw disk drive command
        is entered in the prompt that is displayed.

        Any valid command may be invoked. Some of the most useful commands are:

     \begin{itemize} 
            \item "C:dst=src", copy src file to dst file
            \item "R:dst=src", rename src file to dst file
            \item "S:filename", delete filename
            \item "CD:dirname", change current directory
            \item "MD:dirname", create directory
            \item "RD:dirname", remove directory
    \end{itemize}
    
        Just be careful! There is nothing stopping you from deleting files or even formatting the
        disk.

\section{Miscellaneous functions}

    \subsection{Color settings}
        Both the background and the text color may be changed while using the editor. 
        The program runs in 16 color text mode, so there are 16 background and 16 text colors to choose from.

        Ctrl+T cycles through text color options.
        And Ctrl+B cycles through background colors.

    \subsection{Text buffer size}
        The text buffer is stored in banked RAM, which is 512 kB expandable to 2 MB.

        The text buffer may not exceed the available banked RAM.

        Press Ctrl+M to get information on the remaining memory reserved for the
        text buffer. It will be reported as blocks free. A block may at most hold 251 characters.
    
    \subsection{Programming toolbox}
        Ctrl+F opens the Programming toolbox.

        Currently only one tool, BASLOAD, is supported. By pressing Ctrl+B or only B when
        the toolbox is active, the open text text file is run through BASLOAD. 
        BASLOAD loads and tokenizes a BASIC program stored in a plain text file so that it 
        can run on the X16. The text file needs to saved to disk before the function is used.

    \subsection{Print hard copy}
        Printing to hard copy is supported through loadable drivers. The driver specification
        is available in Appendix C.

        Press Ctrl+H to open the printing dialog. The editor attempts to load the driver from
        the root of the SD card (device 8) using the file name pattern "X16EDITPD-*". It is not
        recommended to store more than one driver in the root folder that matches this
        file name pattern, as you can't affect which driver the editor loads. If you
        want to keep multiple drivers in the root folder you may prepend
        the file name of unused drivers with any character to avoid them being loaded.

        The "printer driver load error" message is displayed if
        \begin{itemize}
            \item no SD card is present in device \#8,
            \item a storage device mounted as device \#8 doesn't support the Kernal's MACPTR function; would require a customized Kernal,
            \item no file matching the driver name pattern is found in the root folder,
            \item any other read error occured while loading the driver, or
            \item the end of the printer driver is not at address \$9eff.
        \end{itemize}

        The "unsupported printer driver" message is displayed if the API version implemented
        by the driver is not supported by the editor. Currently there is only one API version (1).

        What printing options are available is controlled by the driver. Options may be of the
        following types:
        \begin{itemize}
            \item int8
            \item int16
            \item list
            \item string
        \end{itemize}

        Use the up and down arrow keys to highlight the option you want to set. 
        
        Options of types int8, int16 and list can be incremented or decremented using the left 
        and right arrow keys.

        Options of types int8, int16 and string can also be set by pressing the Enter key.
        An input prompt is opened in the status bar, enabling the user to enter a
        new value.

        The driver is responsible for validating option values and to return plain text
        error messages in case of invalid options.

        Press P or Ctrl+P to start printing. The print operation can be paused by
        pressing ESC. The driver can also pause printing, for instance at the end of a page.
        Resume printing by pressing P or Ctrl+P.

        Press S or Ctrl+S to save the current settings as default, if the function is 
        supported by the driver.

        After the printing has finished a "printing complete" message is displayed in
        the status bar.

        You can leave the printing dialog by pressing the ESC key.

    \subsection{Custom key scancode handlers}
        The X16 Kernal allows custom key scancode handlers by changing the vector
        at \$032e-032f. A scancode handler gets the raw IBM key number read from
        the System Management Controller (SMC). The handler can intercept and change
        the scancode before it is processed by the Kernal's keyboard functions.

        The editor itself uses a scancode handler to keep track of modifier key status. The
        handler also consumes all Ctrl key events. By default, the editor's scancode
        handler returns to the Kernal when done, disabling any custom scancode handler that
        was set up before starting the editor.

        You can enable a custom scancode handler by pressing Ctrl+semicolon (;). The editor's
        scancode handler is called first, and the custom scancode handler is linked in
        thereafter. The chain of scancode handlers must eventually return to the Kernal for
        the keyboard to work. This is not within the editor's control after you enable 
        custom scancode handlers.

    \subsection{User-configurable key bindings}
        The shortcut key bindings in X16 Edit are user-configurable.

        On startup the editor reads custom key bindings from the file X16EDITRC in the root directory
        of the SD card. If that file is not found, the default key bindings are
        used.
        
        X16EDITRC is simply a stream of bytes representing the custom shortcuts without any metadata. 
        Each key is represented by the value returned from the Kernal function GETIN.
        
        The values in X16EDITRC are bound to shortcuts in a fixed order, the same order as
        the shortcuts appear in Appendix A.
        
        If X16EDITRC holds fewer values than there are shortcuts, the editor will use
        default bindings for the remaining ones. If there are more values in the file than there
        are shortcuts, the excess is ignored.
        
        To make it a bit easier to setup X16EDITRC, you may use the provided tool
        (X16EDIT-KEYBINDINGS.PRG).

\section{Advanced topics}

    \subsection{Building X16 Edit from source}
        If you like to build X16 Edit yourself you may download the source code from
        www.github.com/stefan-b-jakobsson/x16-edit.

        You will need the cc65 development tools to build the project. You will also need the lzsa compression utility.

        The project is built using Makefile.

        \begin{itemize}
            \item "make" or "make ram" will build the standard RAM version
            \item "make rom" builds the ROM version
            \item "make all" builds both targets
        \end{itemize}

    \subsection{X16 Edit ROM version}
        The ROM version of the editor consists of two consequtive 16 kB ROM banks.

        As of Kernal version R45, the editor is stored in ROM. If you want to store
        your own version of the editor in ROM as well you first need to prepare a custom ROM image. When you
        download the emulator you get the standard rom.bin image. Append
        X16 Edit to the end of it with the following command (Linux and MacOS):

\begin{verbatim}
    cat rom.bin x16edit-rom-x.x.x.bin > customrom.bin
\end{verbatim}

        You then need to write the custom ROM image to the actual ROM circuit to use it on
        real hardware.

        A custom ROM image can easily be attached to the emulator with
        the "-rom" option, for example as follows:

\begin{verbatim}
    x16emu -sdcard sdcard.img -rom customrom.bin
\end{verbatim}

    The Kernal is in development, and the ROM bank layout may change over time. A user
    may also store other utilities in ROM. In order to programatically find in which
    ROM bank X16 Edit is stored, there is an application signature at \$FFF0. The
    signature consists of the text "X16EDIT" followed by three bytes representing
    the program version (major, minor, patch).
        
    If the editor's first ROM bank is 16, it can be started from BASIC like this:

\begin{verbatim}
    BANK 16,16
    SYS $C000
\end{verbatim}

        If you like to use one of the other entry points that are described in Appendix B,
        you need to write a startup program stored in RAM. Code samples doing this
        are also found in Appendix B.

    \subsection{The X16 Edit API}
        The X16 Edit API consists of the following four entry points:

        \begin{itemize}
            \item Default entry point, starts the editor with default options and an empty new text buffer
            \item Load file entry point, starts the editor with default options and then loads the specified text file
            \item Load file with options 1, starts the editor with custom options and then loads the specified text file
            \item Load file with options 2, same as previous, in addition supports moving the cursor to a specified line in the opened file
        \end{itemize}

        Information on how to call the different entry points and code samples is available in Appendix B. 

\appendix
\newpage
\section{List of keyboard shortcuts}
    This is a complete list of keyboard shortcuts supported
    by X16 Edit. You may select commands in any of the following ways:

    \begin{itemize}
        \item Ctrl+shortcut key
        \item Press and release ESC+shortcut key
        \item A function key from the F-key column
    \end{itemize}

    \begin{longtable}[l]{c c l}
        \textbf{Key} & \textbf{F-key} & \textbf{Description} \\
        \hline \\
        G    & F1    & Display built-in help screen \\
        X    & F2    & Exit from the program \\
        O    & F3    & Write text buffer to file \\
        R    & F5    & Open and read a file into the buffer \\
        N    & ---   & Create new text buffer \\
        J    & F4    & Justify text buffer \\
        Y    & F7    & Page up \\
        V    & F8    & Page down \\
        K    & ---   & Cut line or selection to clipboard \\
        C    & ---   & Copy line or selection to clipboard \\
        U    & ---   & Paste (uncut) all content from clipboard \\
        DEL  & ---   & Deletes current line, no copy to clipboard \\
        W    & F6    & Search and find string in buffer (case sensitive) \\
        S    & ---   & Replace string (case sensitive) \\
        L    & ---   & Goto line number \\
        A    & ---   & Toggle auto indent on/off \\
        Z    & ---   & Toggle word wrap on/off \\
        E    & ---   & Change charset \\
        Q    & ---   & Quote mode \\
        I    & ---   & Invoke DOS command \\
        D    & ---   & Set file storage device number, default is 8 \\
        T    &  ---  & Cycle through text colors \\
        B    &  ---  & Cycle through background colors \\
        M    &  ---  & Show memory usage (1 block=251 bytes) \\
        space& ---   & Insert non-breaking space \\
        1..9 & ---   & Set tab stop width \\
        ENTER & ---  & Set line break encoding \\
        F    & ---   & Programming toolbox \\
        ;    & ---   & Custom scancode handler on/off \\
        H    & ---   & Print hard copy \\
    \end{longtable}

\newpage
\section{X16 Edit API}

    \subsection{Default entry point}

        \textbf{Purpose:} Start the editor with default options and an empty new text buffer

        \textbf{Call address:} \$080D (RAM version), \$C000 (ROM version)
    
        \textbf{Parameters:} None

    \subsection{Load file entry point}
        
        \textbf{Purpose:} Start the editor and load a specified text file

        \textbf{Call address:} \$0810 (RAM version), \$C003 (ROM version)

        \textbf{Parameters:} 

        \begin{longtable}[l]{l l l}
            \textbf{Register} & \textbf{Address} & \textbf{Description} \\
	        \hline \\
            X   &          & First RAM bank used by the program \\
            Y   &          & Last RAM bank used by the program \\
            r0  & \$02--03 & Pointer to file name \\
            r1L & \$04     & File name length, or 0=no file \\
        \end{longtable}

        If the specified file does not exist, the editor will display an
        error message. If the file name length is 0, the program will not try
        to load a text file on startup.

        The first and last RAM bank settings control what part of banked RAM is
        used by the program. This option may be used to reserve the rest of banked RAM
        for other purposes.

    \subsection{Load file with options entry point 1}

        \textbf{Purpose:} Start editor with custom options and then load the specified text file

        \textbf{Call address:} \$0813 (RAM version), \$C006 (ROM version)

        \textbf{Parameters:}

        \begin{longtable}[l]{l l l l}
	        \textbf{Register} & \textbf{Address} & \textbf{Bits} & \textbf{Description} \\
	        \hline \\
            X   &          &      & First RAM bank used by the program \\
            Y   &          &      & Last RAM bank used by the program \\
            r0  & \$02--03 &      & Pointer to file name \\
            r1L & \$04     &      & File name length, or 0=no file \\
            r1H & \$05     & 0    & Auto indent on/off \\
            r1H & \$05     & 1    & Word wrap on/off \\
            r1H & \$05     & 2--7 & Unused \\
            r2L & \$06     &      & Tab stop width (1..9) \\
            r2H & \$07     &      & Word wrap position (10..250) \\
            r3L & \$08     &      & Current device number (8..30) \\
            r3H & \$09     & 0--3 & Text color \\
            r3H & \$09     & 4--7 & Background color \\
            r4L & \$0A     & 0--3 & Header text color \\
            r4L & \$0A     & 4--7 & Header background color \\
            r4H & \$0B     & 0--3 & Status bar text color \\
            r4H & \$0B     & 4--7 & Status bar background color \\
        \end{longtable}

        Parameters out of range are silently ignored, and default
        values are used instead.

        Color settings are ignored if both the text and background color is 0.

        If the specified file does not exist, the editor will display an
        error message. If the file name length is 0, the program will not try
        to load a text file.

        The first and last RAM bank settings control what part of banked RAM is
        used by the program. This option may be used to reserve the rest of banked RAM
        for other purposes.

    \subsection{Load file with options entry point 2}

        \textbf{Purpose:} Start editor with custom options and then load the specified text file. In addition to Load file with options entry point 1, this entry
        point adds an optional line number whereto the cursor is moved after opening the file.

        \textbf{Call address:} \$0816 (RAM version), \$C009 (ROM version)

        \textbf{Parameters:}

        \begin{longtable}[l]{l l l l}
	        \textbf{Register} & \textbf{Address} & \textbf{Bits} & \textbf{Description} \\
	        \hline \\
            X   &          &      & First RAM bank used by the program \\
            Y   &          &      & Last RAM bank used by the program \\
            r0  & \$02--03 &      & Pointer to file name \\
            r1L & \$04     &      & File name length, or 0=no file \\
            r1H & \$05     & 0    & Auto indent on/off \\
            r1H & \$05     & 1    & Word wrap on/off \\
            r1H & \$05     & 2--7 & Unused \\
            r2L & \$06     &      & Tab stop width (1..9) \\
            r2H & \$07     &      & Word wrap position (10..250) \\
            r3L & \$08     &      & Current device number (8..30) \\
            r3H & \$09     & 0--3 & Text color \\
            r3H & \$09     & 4--7 & Background color \\
            r4L & \$0A     & 0--3 & Header text color \\
            r4L & \$0A     & 4--7 & Header background color \\
            r4H & \$0B     & 0--3 & Status bar text color \\
            r4H & \$0B     & 4--7 & Status bar background color \\
            r5--r6L & \$0C   &      & Goto line number (24 bits) \\
        \end{longtable}

        Parameters out of range are silently ignored, and default
        values are used instead.

        Color settings are ignored if both the text and background color is 0.

        If the specified file does not exist, the editor will display an
        error message. If the file name length is 0, the program will not try
        to load a text file.

        The first and last RAM bank settings control what part of banked RAM is
        used by the program. This option may be used to reserve the rest of banked RAM
        for other purposes.

        The cursor is moved to the goto line number (24 bits) on file open. The goto
        line number is ignored if 0. The cursor is placed on the last line, if
        the requested line does not exist.

    \subsection{Code samples for the RAM version}

    \subsubsection{Default entry point}
\begin{verbatim}
    jsr $080d  ; No parameters, just call the entry point
    rts
\end{verbatim}

    \subsubsection{Load file entry point}
\begin{verbatim}
    ldx #$01               ; First RAM bank used by the editor
    ldy #$ff               ; And last RAM bank
    lda #<fname            ; Pointer to file name (LSB)
    sta $02                ; Store in r0L
    lda #>fname            ; Pointer to file name (MSB)
    sta $03                ; Store in r0H
    lda #fname_end-fname   ; File name length
    sta $04                ; Store in r1L
    jsr $0810              ; Call entry point
    rts

    fname:
        .byt "mytextfile.txt"
    fname_end:
\end{verbatim}

\subsubsection{Load file with options entry point 1}
\begin{verbatim}
    ldx #$01               ; First RAM bank used by the editor
    ldy #$ff               ; And last RAM bank
    lda #<fname            ; Pointer to file name (LSB)
    sta $02                ; Store in r0L
    lda #>fname            ; Pointer to file name (MSB)
    sta $03                ; Store in r0H
    lda #fname_end-fname   ; File name length
    sta $04                ; Store in r1L
    lda #$01               ; Auto-indent on, word wrap off
    sta $05                ; Store in r1H
    lda #$04               ; Tab width
    sta $06                ; Store in r2L
    lda #$28               ; Word wrap position
    sta $07                ; Store in r2H
    lda #$08               ; Disk drive device number
    sta $08                ; Store in r3L
    lda #$b1               ; Text white, background light green
    sta $09                ; Store in r3H (screen color)
    lda #$07               ; Text yellow, background black
    sta $0a                ; Store in r4L (header color)
    lda #$00               ; Use default color
    sta $0b                ; Store in r4H (status bar color)
    jsr $0813              ; Call entry point
    rts

fname:
    .byt "mytextfile.txt"
fname_end:
\end{verbatim}

\subsubsection{Load file with options entry point 2}
\begin{verbatim}
    ldx #$01               ; First RAM bank used by the editor
    ldy #$ff               ; And last RAM bank
    lda #<fname            ; Pointer to file name (LSB)
    sta $02                ; Store in r0L
    lda #>fname            ; Pointer to file name (MSB)
    sta $03                ; Store in r0H
    lda #fname_end-fname   ; File name length
    sta $04                ; Store in r1L
    lda #$01               ; Auto-indent on, word wrap off
    sta $05                ; Store in r1H
    lda #$04               ; Tab width
    sta $06                ; Store in r2L
    lda #$28               ; Word wrap position
    sta $07                ; Store in r2H
    lda #$08               ; Disk drive device number
    sta $08                ; Store in r3L
    lda #$b1               ; Text white, background light green
    sta $09                ; Store in r3H (screen color)
    lda #$07               ; Text yellow, background black
    sta $0a                ; Store in r4L (header color)
    lda #$00               ; Use default color
    sta $0b                ; Store in r4H (status bar color)
    lda #$2c               ; Goto line number 300 on file open
    sta $0c
    lda #$01
    sta $0d
    lda #$00
    sta $0e
    jsr $0816              ; Call entry point
    rts

fname:
    .byt "mytextfile.txt"
fname_end:
\end{verbatim}

    \subsection{Code samples for the ROM version}
    
    \subsubsection{Search ROM banks for X16 Edit}

    The X16 Edit ROM bank may be identified by the
    application signature ("X16EDIT") stored at \$FFF0. This
    code sample searches all ROM banks for the signature.
    If found, the ROM bank is returned in A with carry clear; otherwise
    carry is set on return.

\begin{verbatim}
find_me:
    lda $01                ; Store current ROM bank on stack
    pha
    stz $01                ; Prepare searching from ROM bank 0
    ldy #$00

scan:
    lda $fff0,y            ; Signature starts at $fff0
    cmp signature,y
    bne next               ; Signature didn't match, check next ROM bank
    iny                    ; Increase char pointer
    cpy #$07               ; Have we got 7 matching chars? If not, keep looking
    bne scan
    clc                    ; Set C = 0 as indicator X16 Edit was found
    lda $01                ; Load ROM bank into A
    bra exit

next:
    ldy #$00               ; Reset char pointer
    inc $01                ; Select next ROM bank
    lda $01            
    cmp #$20               ; Have we checked all ROM banks?
    bne scan
    sec                    ; Set C = 1 as indicator X16 Edit was not found

exit:
    plx                    ; Restore original ROM bank
    stx $01
    rts

signature: .byt $58,$31,$36,$45,$44,$49,$54     ; = "X16EDIT"
\end{verbatim}

    \subsubsection{Default entry point}
\begin{verbatim}
    lda $01                ; Store current ROM bank on stack
    pha
    jsr find_me            ; Search ROM banks
    bcs done               ; Exit if X16 Edit wasn't found
    sta $01                ; Set ROM bank
    jsr $c000              ; Call entry point
done:
    pla                    ; Restore original ROM bank
    sta $01
    rts
\end{verbatim}

    \subsubsection{Load file entry point}
\begin{verbatim}
    lda $01                ; Store current ROM bank on stack
    pha
    jsr find_me            ; Search ROM banks
    bcs done               ; Exit if X16 Edit wasn't found
    sta $01                ; Set ROM bank  
    ldx #$01               ; First RAM bank used by the editor
    ldy #$ff               ; And last RAM bank
    lda #<fname            ; Pointer to file name (LSB)
    sta $02                ; Store in r0L
    lda #>fname            ; Pointer to file name (MSB)
    sta $03                ; Store in r0H
    lda #fname_end-fname   ; File name length
    sta $04                ; Store in r1L
    jsr $c003              ; Call entry point
done:
    pla                    ; Restore original ROM bank
    sta $01
    rts

fname:
    .byt "mytextfile.txt"
fname_end:
\end{verbatim}

    \subsubsection{Load file with options entry point 1}
\begin{verbatim}
    lda $01                ; Store current ROM bank on stack
    pha
    jsr find_me            ; Search ROM banks
    bcs done               ; Exit if X16 Edit wasn't found
    sta $01                ; Set ROM bank
    
    ldx #$01               ; First RAM bank used by the editor
    ldy #$ff               ; And last RAM bank
    lda #<fname            ; Pointer to file name (LSB)
    sta $02                ; Store in r0L
    lda #>fname            ; Pointer to file name (MSB)
    sta $03                ; Store in r0H
    lda #fname_end-fname   ; File name length
    sta $04                ; Store in r1L
    lda #$01               ; Auto-indent on, word wrap off
    sta $05                ; Store in r1H
    lda #$04               ; Tab width
    sta $06                ; Store in r2L
    lda #$28               ; Word wrap position
    sta $07                ; Store in r2H
    lda #$08               ; Disk drive device number
    sta $08                ; Store in r3L
    lda #$b1               ; Text white, background light green
    sta $09                ; Store in r3H (screen color)
    lda #$07               ; Text yellow, background black
    sta $0a                ; Store in r4L (header color)
    lda #$00               ; Use default color
    sta $0b                ; Store in r4H (status bar color)
    jsr $c006              ; Call entry point
done:
    pla                    ; Restore original ROM bank
    sta $01
    rts
   
fname:
    .byt "mytextfile.txt"
fname_end:
\end{verbatim}


    \subsubsection{Load file with options entry point 2}
\begin{verbatim}
    lda $01                ; Store current ROM bank on stack
    pha
    jsr find_me            ; Search ROM banks
    bcs done               ; Exit if X16 Edit wasn't found
    sta $01                ; Set ROM bank
    
    ldx #$01               ; First RAM bank used by the editor
    ldy #$ff               ; And last RAM bank
    lda #<fname            ; Pointer to file name (LSB)
    sta $02                ; Store in r0L
    lda #>fname            ; Pointer to file name (MSB)
    sta $03                ; Store in r0H
    lda #fname_end-fname   ; File name length
    sta $04                ; Store in r1L
    lda #$01               ; Auto-indent on, word wrap off
    sta $05                ; Store in r1H
    lda #$04               ; Tab width
    sta $06                ; Store in r2L
    lda #$28               ; Word wrap position
    sta $07                ; Store in r2H
    lda #$08               ; Disk drive device number
    sta $08                ; Store in r3L
    lda #$b1               ; Text white, background light green
    sta $09                ; Store in r3H (screen color)
    lda #$07               ; Text yellow, background black
    sta $0a                ; Store in r4L (header color)
    lda #$00               ; Use default color
    sta $0b                ; Store in r4H (status bar color)
    lda #$2c               ; Goto line 300 on file open
    sta $0c
    lda #$01
    sta $0d
    lda #$00
    sta $0e
    jsr $c009              ; Call entry point
done:
    pla                    ; Restore original ROM bank
    sta $01
    rts
   
fname:
    .byt "mytextfile.txt"
fname_end:
\end{verbatim}

\newpage
\section{Printer driver specification}

    \subsection{Introduction}

        Built-in support for the vast amount of printers that exist and all the ways printers
        can be connected to the system would be difficult at best. To overcome this complexity
        and to enable users to use any setup they may have, printing is supported through
        loadable drivers.

        A printer driver is a separate machine language program implementing the API
        specified below.

        The driver must be stored in the root folder of the SD card. The name of the file
        must begin with "X16EDITPD-", followed by an optional name identifying the printer,
        for instance "X16EDITPD-MPS801". You must be able to load the driver into the
        correct memory address using the BASIC command LOAD "X16EDITPD-*",8,1.

        The actual printing is done by three functions: channel\_open, print\_char, and
        channel\_close. The other functions are mainly used to initialize the driver, and to
        get and set printer options.

        Options may be any of the following types:

        \begin{itemize}
            \item int8: an 8 bit integer value
            \item int16: a 16 bit integer value
            \item list: represented by an array of null-terminated string
            \item string: a null-terminated string
        \end{itemize}

    \subsection{Driver API version 1}

        The driver may start at any point in low RAM but must always end at address \$9eff.
        If you want the driver to work with the RAM based version of X16 Edit, it must, however, 
        avoid affecting the range of low RAM used by the editor program.

        The zero page addresses \$36-39 are reserved for and can be used by the printer driver.
        The driver should not use any of the golden RAM (\$0400-\$7fff) or banked
        RAM (\$a000-\$bfff). Using any other parts of zero page, golden RAM or
        banked RAM may cause conflicts with the editor. 
        
        The driver must implement the following jump table. Each entry is expected to be a JMP instruction.

        \begin{longtable}[l]{l l}
            \textbf{Address} & \textbf{Name} \\
            \hline \\
            \$9efd & init \\
            \$9efa & get\_api\_version \\
            \$9ef7 & get\_driver\_name \\
            \$9ef4 & set\_charset \\
            \$9ef1 & get\_option\_count \\
            \$9eee & get\_option\_labels \\
            \$9eeb & get\_option\_value \\
            \$9ee8 & set\_set\_option\_value \\
            \$9ee5 & option\_add\_offset \\
            \$9ee2 & channel\_open \\
            \$9edf & print\_char \\
            \$9edc & channel\_close \\            
            \$9ed9 & get\_message \\            
            \$9ed6 & save\_defaults \\
        \end{longtable}

        All string messages returned by the driver must be unshifted PETSCII
        to display correctly in the print dialog.

        \subsubsection{init}
            Purpose: Called once by the editor after loading the
            driver into memory. Can be used to make necessary
            setup in the driver before any other functions are
            called.  
            
            Call address: \$9efd.

            Input: None.
            
            Output: None.
            
            Error returns: None.
            
            Registers affected: May affect all registers.

        \subsubsection{get\_api\_version}
            Purpose: Returns the printer driver API version that
            the driver was designed for.

            Call address: \$9efa.

            Input: None.
            
            Output: .A = API version number.

            Error returns: None.

            Registers affected: .A
        
        \subsubsection{get\_driver\_name}
            Purpose: Returns the driver name, typically the
            name of the supported printer.

            Call address: \$9ef7.

            Input: None.

            Output: .X/.Y = Pointer to null-terminated string.

            Error returns: None.

            Registers affected: May affect all registers.

        \subsubsection{set\_charset}
            Purpose: Sets the charset used to display the current
            document. Called by the editor just before opening
            the printer channel.

            Call address: \$9ef4.

            Input: .A = Charset (0 = ASCII, 1 = PETSCII upper case, 2 = PETSCII lower case).

            Output: None.

            Error returns: .C = 1 if invalid input.

            Registers affected: May affect all registers.

        \subsubsection{get\_option\_count}
            Purpose: Returns the number of options that may be changed through the
            editor's print dialog.

            Call address: \$9ef1.

            Input: None.

            Output: .A = Option count.

            Error returns: None.

            Registers affected: .A.

        \subsubsection{get\_option\_labels}
            Purpose: Returns a pointer to a null-terminated string that
            contains labels for all driver options. Labels must be
            separated from each other with CR (PETSCII 13). The number
            of labels must match the option count returned by the
            function get\_option\_count.

            Call address: \$9eee.

            Input: None.

            Output: .X/.Y = Pointer to null-terminated string.

            Error returns: None.

            Registers affected: May affect all registers.

            Example return value, from the MPS-801 driver:
\begin{verbatim}
    SERIAL DEVICE:<CR>
    PAGE HEIGHT..:<CR>
    TOP MARGIN...:<CR>
    BOTTOM MARGIN:<CR>
    ON PAGE BREAK:<NULL>
\end{verbatim}

            It is recommended that all lines have the same length and
            that each line ends with a colon, as demonstrated in the
            above example. The editor displays all option values starting
            from the column after the longest label.

        \subsubsection{get\_option\_value}
            Purpose: Returns the type and value of the specified driver option.

            Call address: \$9eeb.

            Input: .X = Option index (zero-based).

            Output: .A = Type (0 = uint8, 1 = uint16, 2 = list item, 3 = null-terminated string). 
            .X = uint8 value.
            .X/.Y = uint16 value.
            .X/.Y = Pointer to null-terminated string representing a list item or a string.

            Error returns: .C = 1 if option does not exist. A plain text error message is 
            expected to be available through the get\_message function.

            Registers affected: May affect all registers.

        \subsubsection{set\_option\_value}
            Purpose: Changes the specified driver option. The new option value
            is always received as a null-terminated string. The driver is
            responsible for converting and verifying the value.

            Call address: \$9ee8.

            Input: .X = Option index. .A/.Y = Pointer to null-terminated string. This
            function always expects values in string format. It is for the driver
            to convert the string into format usable by the driver.

            Output: .A = Type (0 = uint8, 1 = uint16, 2 = list item, 3 = null-terminated string). 
            .X = uint8 value.
            .X/.Y = uint16 value.
            .X/.Y = Pointer to null-terminated string representing a list item or a string. 

            Error returns: .C = 1 if option does not exist or if the value is
            invalid. A plain text error message is expected to be available through the
            get\_message function.

            Registers affected: May affect all registers.

        \subsubsection{option\_add\_offset}
            Purpose: Adds a signed 8-bit integer to the selected option. The option
            must be any of the following types: int8, int16 or list.
            
            Call address: \$9EE5.

            Input: .A = offset to add.

            Output: .A = Type (0 = uint8, 1 = uint16, 2 = list item, 3 = null-terminated string). 
            .X = uint8 value.
            .X/.Y = uint16 value.
            .X/.Y = Pointer to null-terminated string representing a list item or a string. 

            Error returns: .C = 1 if option does not exist. A plain text error message is expected 
            to be available through the get\_message function.

            Registers affected: May affect all registers.

        \subsubsection{channel\_open}
            Purpose: Opens the printer channel. This function may also do any other
            necessary tasks to prepare for printing, for instance setting printer
            options or resetting the printer.

            Call address: \$9ee2.

            Input: None.

            Output: None.

            Error returns: .C = 1 if an error occured. A plain text error message is expected 
            to be available through the get\_message function.

            Registers affected: May affect all registers.

        \subsubsection{print\_char}
            Purpose: Prints a character. The driver may pause printing before the
            received character is outputted, indicated by returning .A = 1. This option may 
            for instance be used to pause printing on page breaks. The text editor will resend 
            the same character when printing is resumed.

            Call address: \$9edf.

            Input: .A = character, encoded as set by the set\_charset function.

            Output: .A = response code (0 = OK, 1 = Printing paused, need to resend the same
            character).

            Error returns: .C = 1 if an error occured. A plain text error message is expected
            to be available through the get\_message function.

            Registers affected: .X, .Y

        \subsubsection{channel\_close}
            Purpose: Closes the printer channel. This function may also do any other tasks
            needed to finish printing.

            Call address: \$9edc.

            Input: None.

            Output: None.

            Error returns: .C = 1 on error. A plain text error message is expected
            to be available through the get\_message function.

            Registers affected: May affect all registers.

        \subsubsection{get\_message}
            Purpose: Returns a pointer to the last driver (error) message. The
            message is cleared after calling this function. If no message is
            available, this function should return a pointer to  a zero 
            length string.

            Call address: \$9ed9.

            Input: None.

            Output: .X/.Y = Pointer to null-terminated string.

            Error returns: None.

            Registers affected: May affect all registers.

        \subsubsection{save\_defaults}
            Purpose: Saves default printer options in a way that can be
            retrieved when loading or initializing the driver next time.
            The driver can load default options, for instance as part
            of the init function.

            Call address: \$9ed6.

            Input: None.

            Output: None.

            Error returns: .C = 1 on error. A plain text error message is expected
            to be available through the get\_message function.

            Registers affected: May affect all registers.

\newpage 
\section{License}
	Copyright 2020--2026, Stefan Jakobsson.

	The X16 Edit program, including this manual, is released under the 2-Clause BSD
    License.
    
    Redistribution and use in source and binary forms, with or without modification, 
    are permitted provided that the following conditions are met:

    1. Redistributions of source code must retain the above copyright notice, this 
       list of conditions and the following disclaimer.

    2. Redistributions in binary form must reproduce the above copyright notice, 
       this list of conditions and the following disclaimer in the documentation 
       and/or other materials provided with the distribution.

    THE SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS “AS IS” 
    AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE 
    IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE 
    DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE 
    FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL 
    DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR 
    SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER 
    CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, 
    OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE 
    OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.


\end{document}
